### 2025-09-06 base-rates.qmd
20250906 - overconfidence bias

@@

We examine the base rates of rookie performance compared to performance of non-rookies.
Rookies who play tend to be better players who were drafted in early rounds of the National Football League (NFL) draft.
Thus, to more fairly compare rookies and non-rookies, we examine only players who were drafted in the first two rounds of the NFL Draft.

>>>>

Teams in the NFL also tend to neglect base rates.
For instance, NFL teams give too much weight to scouting evidence in deciding which players to draft, and fail to integrate such evidence with the prior probabilities of a player's successful future performance [i.e., the base rate of success; @Massey2013].
"...consider how suspiciously often we hear a college prospect described as a 'once-in-a-lifetime player'." [@Massey2013, p. 1482].

@@

We examine the base rates of rookie performance compared to performance of non-rookies.
It can be challenging to predict rookie performance because they have not yet played in the NFL.
Thus, it is important to consider the prior probabilities (base rates) of success among rookies in the NFL.
Rookies who play tend to be better players who were drafted in early rounds of the National Football League (NFL) draft.
Thus, to more fairly compare rookies and non-rookies, we examine only players who were drafted in the first two rounds of the NFL Draft.

### 2025-09-06 cognitive-bias.qmd
20250906 - overconfidence bias

*Overestimation* involves believing that one will perform better than one actually performs.
Overestimation can be identified with a [calibration plot](#sec-calibrationPlot) of the predicted performance versus actual performance, where the person's predicted performance is systematically higher (in at least some cases) than their actual performance.
Overestimation corresponds to the "[overprediction](#sec-calibration)" form of [miscalibration](#sec-calibration).

@@

The confidence that people tend to have in their predictions depends less and the quantity and quality of the evidence for their predictions and more on the quality of the narrative they can tell about the (little) information they have immediate access to [@Kahneman2011].
For instance, people tend not to consider other relevant, important factors that they do not have immediate access to, consistent with the [WYSIATI heuristic](#sec-heuristicsWYSIATI).
Instead, people tend to focus on the coherence and plausibility of the story, which differs from its probability [@Kahneman2011].
That is, one's confidence is not necessarily related to the likelihood that the judgment is correct—it often reflects the "illusion of validity" [@Kahneman2011].
The illusion of validity is the unwarranted confidence that arises from a perceived good fit between the predicted outcome and the input information [@Tversky1974].
As one gains more knowledge in a domain, the person also tends to become unrealistically overconfident in their ability to make accurate judgments and predictions [@Kahneman2011].
In general, people tend to focus on what they know and to neglect what they do not know leading to overconfidence in one's beliefs [@Kahneman2011].

>>>>

*Overestimation* involves believing that one will perform better than one actually performs.
It thus reflects a greater mean confidence than overall accuracy [@Kahneman1996].
In general, people tend to show excessive confidence in their beliefs; for instance, when people express 90% confidence about their belief, they may be correct only about 75% of the time [@Kahneman1996].
As an example of overconfidence in football, early picks in the NFL draft tend to be considerably overvalued because teams believe their predictions will be more accurate than they actually are [@Massey2013].
The probability that a player is better than the next player chosen in the draft at the same position is ~52%, or basically a coin flip [@Massey2013].
As noted by @Massey2013 (p. ) "...decision makers often know less than they think they know...The problem is not that future performance is difficult to predict, but that decision makers do not appreciate how difficult it is."
This suggests that a) the presence of high stakes for the prediction does not eliminate overconfidence, and that b) teams should "trade down" to acquire more, later picks if they possess very early picks [@Massey2013].
Overestimation can be identified with a [calibration plot](#sec-calibrationPlot) of the predicted performance versus actual performance, where the person's predicted performance is systematically higher (in at least some cases) than their actual performance.
Overestimation corresponds to the "[overprediction](#sec-calibration)" form of [miscalibration](#sec-calibration).

@@

The confidence that people tend to have in their predictions depends less and the quantity and quality of the evidence for their predictions and more on the quality of the narrative they can tell about the (little) information they have immediate access to [@Kahneman2011].
For instance, people tend not to consider other relevant, important factors that they do not have immediate access to, consistent with the [WYSIATI heuristic](#sec-heuristicsWYSIATI).
Instead, people tend to focus on the coherence and plausibility of the story, which differs from its probability [@Kahneman2011].
That is, one's confidence is not necessarily related to the likelihood that the judgment is correct—it often reflects the "illusion of validity" [@Kahneman2011].
The illusion of validity is the unwarranted confidence that arises from a perceived good fit between the predicted outcome and the input information [@Tversky1974].
As one gains more knowledge in a domain, the person also tends to become unrealistically overconfident in their ability to make accurate judgments and predictions [@Kahneman2011].
In general, people tend to focus on what they know and to neglect what they do not know leading to overconfidence in one's beliefs [@Kahneman2011].
Moreover, overconfidence tends to be exacerbated by information—the more information people have, the more they tend to be overconfident [@Massey2013].
Thus, lots of scouting evidence may lead teams and fantasy managers to be overconfident in the draft.

### 2025-09-06 sports-cognitive-psychology.qmd
20250906 - overconfidence bias

Examples of changes in football due to the use of analytics include more often going for it on fourth down, a greater emphasis on the passing game, drafting Running Backs later in the draft (and, more generally, valuing Running Backs less), and trading down in the draft to obtain more low picks (rather than having fewer high picks) because top picks are frequently overvalued [@Massey2013].

>>>>

Examples of changes in football due to the use of analytics include more often going for it on fourth down, a greater emphasis on the passing game, drafting Running Backs later in the draft (and, more generally, valuing Running Backs less), and trading down in the draft to obtain more low picks (rather than having fewer high picks) because top picks are frequently overvalued [@Massey2013].
Although both performance and compensation decline with draft order in the NFL, compensation declines more steeply than performance [@Massey2013].
The highest value picks—in terms of the greatest expected performance relative to compensation (i.e., performance per dollar)—are near the beginning of the second round of the NFL draft [@Massey2013].
According to their analysis, @Massey2013 found that the least valuable pick in the first round of the NFL draft (based on expected performance relative to compensation) is the 1st overall pick!
Thus, they refer to the first pick as the "loser's curse" [@Massey2013].
In general, the overvaluation of picks tends to be greatest at the top of the draft, even accounting for the possibility of drafting superstars [@Massey2013].
